A formalization of semi-simplicial and semi-cubical sets in dependent type theory

### 2023-01-11 paper/paper.tex
A polishing step

\begin{document}
\maketitle
\begin{abstract}
  We present a parametricity-based construction of semi-simplicial sets and semi-cubical sets in indexed form, formulated in dependent type theory. The accompanying artifact is an axiom-less~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to $\nu$-types; if $\nu$ is finite (say, unit type corresponding to augmented semi-simplicial types), this requirement vanishes.} Coq formalization that uses sophisticated rewriting techniques and pushes the boundaries of proof assistant technology.
\end{abstract}

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds72} interprets types as relations characterizing the observational equality of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has a cubical flavor~\cite{johann17,altenkirch15,moulin16}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates instead of relations, in which case we obtain a form of realizability~\cite{bernardy12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Observation by Hugo Moeneclaey}.

We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both semi-simplicial and semi-cubical sets obtained by applying simple rules iterating parametricity. In contrast to the presheaf presentation of semi-simplicial and semi-cubical sets~\cite{cchm}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of semi-simplices or semi-cubes in dimension $n+1$
\emph{fibered} over the set of semi-simplices or semi-cubes in dimension $n$, we consider families of semi-simplices and semi-cubes indexed over their faces.

The outline of the paper is as follows. First, we briefly describe semi-simplicial and semi-cubical sets, and our generalization to $\nu$-sets. We then proceed to explain our construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning in sets. We end with some of the finer details of our mechanization.

@@

\begin{notation}[Sequences of integers]
  We denote sequences of integers by $\{i_1, \ldots i_n\}$. In particular, the empty sequence is written $\{\}$.
\end{notation}

@@

\begin{definition}[$\DeltaPlus$]
  \begin{figure*}[!t]
    \begin{align*}
      \obj(\DeltaPlus) & := \Nat                                                                     \\
      \hom(p, n)       & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in l = p}\}        \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                      \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure*}

@@

@@

  \begin{align*}
    \DeltaPlus^n    & : \Set_{\DeltaPlus}      \\
    \DeltaPlus^n(p) & := \Hom(p, n)            \\
    \DeltaPlus^n(f) & := \lambda g . g \circ f
  \end{align*}
\end{definition}

@@

\subsection{Semi-cubical sets}
\begin{definition}[\Cube]
  $\Cube$ is defined similarly to $\DeltaPlus$, except that we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$. We use $L$ and $R$ to indicate opposite faces of a cube. The definition is shown in Fig. \ref{fig:cube}.

  \begin{figure*}[!t]
    \begin{align*}
      \obj(\Cube) & := \Nat                                                                                \\
      \hom(p, n)  & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                            \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
    \end{align*}
    \caption{Definition of \Cube}\label{fig:cube}
  \end{figure*}

@@

  \begin{align*}
    \Cube^n    & : \CSet                  \\
    \Cube^n(p) & := \Hom(p, n)            \\
    \Cube^n(f) & := \lambda g . g \circ f
  \end{align*}
\end{definition}

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the straightforward generalization of augmented semi-simplicial sets and semi-cubical sets to an arbitrary cardinal, where morphisms are sequences of elements of an arbitrary set $\nu$, extended with $\kstar$. Then, we observe:

@@

\section{Our construction}
In this section, we gradually build up an intuition for our construction, followed by the full details of our construction.

\subsection{Homotopy type theory}
Homotopy type theory (HoTT) is a logical foundation based on Martin-Löf type theory and the univalence axiom~\cite{hottbook}. HoTT is based on the notion of a basic \Type, and the notion of sets can be recovered as \U, where $\U$ is a straightforward packaging of two fields:

@@

In the $\U$ universe, three properties hold:

\begin{enumerate}
  \item[(i)] UIP holds on the unit type.
  \item[(ii)] UIP propagates to $\Sigma$-types.
  \item[(iii)] UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

$\U$ provides us a form of unicity of identity of proofs (UIP), which is a form of proof irrelevance:

\begin{align*}
  \forall x y, \forall p q : x = y, p = q
\end{align*}

which is to say that any two proofs of \emph{equality} of the same two types are \emph{equal}. In other words, the proofs cannot be distinguished from one another.

An equivalent definition of UIP is given by \textsf{UIP\_refl}:

\begin{align*}
  \textsf{UIP\_refl} & : \forall x, \forall p : x = x, p = \textsf{eq\_refl } x
\end{align*}

In type theory, UIP can be proved on the unit type and bool type:

\begin{align*}
  \textsf{eq\_refl}        & : \forall x, x = x                                                    \\
  \textsf{UIP\_refl\_unit} & : \forall u : \textsf{unit}, \forall x : u = u, x = \textsf{eq\_refl} \\
  \textsf{UIP\_refl\_bool} & : \forall b : \textsf{bool}, \forall x : b = b, x = \textsf{eq\_refl} \\
\end{align*}

\subsection{Fibered versus indexed representation}
It is well-known in set theory that families of small sets can be represented in a fibered or indexed way. In the fibered way, a family over $T$ is represented a pair of a set $S$, and a function $S \rightarrow T$. The function classifies the original set into fibers. In contrast, in the indexed way, a family over $T$ is represented as a functional relation with domain $T$, which itself, by replacement axiom, is a set.

@@

The latter correspondence can be expressed in HoTT, for $T: \Type$, as an equivalence between $\Sigma S: \Type\;.\;S \rightarrow T$ and $T \rightarrow \Type$, which can be proved using univalence. This correspondence also holds on \U, for $T: \U$:

@@

Let us now come back to cubical sets. A cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.

\begin{equation*}
  \begin{tikzcd}
    X_0 : \U & X_1 : \U \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] & X_2 : \U \arrow[l, "\delta^\epsilon_1" description, shift left=6] \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] \arrow[l, "\delta^\omega_1" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

@@

\subsection{Intuition for our formal construction}
\begin{notation}[$\Sigma$]
  In the prose that follows, we assume the existence of $\Sigma$-types with projections $\hd$ and $\tl$.
\end{notation}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Constructions such as semi-simplicial and semi-cubical sets can be defined in the ``usual way'' as presheaves over respectively the semi-simplex or semi-cube category, which we shall call ``fibered'' definitions, but also, as done e.g. in Voevodsky~\cite{} and in previous work~\cite{} as an ``dependently-typed'', or, say, ``indexed'' construction.

This paper describes an alternative indexed construction of both augmented semi-simplicial and semi-cubical sets based on the property that augmented semi-simplicial and semi-cubical sets naturally arise as respectively unary and binary iterated parametricity-based constructions.

Additionally, the axiom-free construction has been fully formalized in Coq's dependent type theory, which turned out not to be so straightforward.
\end{abstract}

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds72} interprets types as relations characterizing the observational equality of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has a cubical flavor~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates instead of relations, in which case we obtain a form of realizability~\cite{bernardy12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Private communication by Thorsten Altenkirch and Hugo Moeneclaey}.

We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both semi-simplicial and semi-cubical sets obtained by applying simple rules iterating parametricity. In contrast to the standard presheaf presentation of semi-simplicial and semi-cubical sets~\cite{fri08,grandis03,buchholtz1}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of semi-simplices or semi-cubes in dimension $n+1$
\emph{fibered} over the set of semi-simplices or semi-cubes in dimension $n$, we consider families of semi-simplices and semi-cubes indexed over their faces.

The outline of the paper is as follows. First, we briefly describe a free-monoid-style (?) in the fibered presentation of semi-simplicial and semi-cubical sets, and our generalization to $\nu$-sets. We then proceed to explain our indexed construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning with sets inside type theory. We end with some of the finer details of our mechanization.

@@

\begin{notation}[Finite sequences]
  We denote finite sequences by $\{i_1, \ldots i_n\}$ for $i_j$ ranging over some domain. In particular, the empty sequence is written $\{\}$.
\end{notation}

@@

\begin{definition}[$\DeltaPlus$]
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\DeltaPlus) & := \Nat                                                                     \\
      \Hom(p, n)       & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}        \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                      \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure*}

@@

To provide examples, we define the standard augmented $n$-semi-simplex.

@@

  \begin{align*}
    \DeltaPlus^n    & : \Set_{\DeltaPlus}      \\
    \DeltaPlus^n(p) & := \Hom(p, n)            \\
    \DeltaPlus^n(f) & := \lambda g .\, g \circ f
  \end{align*}
\end{definition}

@@

\subsection{Semi-cubical sets}

Semi-cubical sets are defined like augmented semi-simplicial sets except that $\DeltaPlus$ is replaced by $\Cube$ in which we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$.

\begin{definition}[\Cube]
  The definition of $\Cube$ is shown in Fig.~\ref{fig:cube}.
  The symbols $L$ and $R$ indicate opposite faces of a cube.
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\Cube) & := \Nat                                                                                \\
      \Hom(p, n)  & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                            \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
    \end{align*}
    \caption{Definition of \Cube}\label{fig:cube}
  \end{figure*}

@@

  \begin{align*}
    \Cube^n    & : \CSet                  \\
    \Cube^n(p) & := \Hom(p, n)            \\
    \Cube^n(f) & := \lambda g .\, g \circ f
  \end{align*}
\end{definition}

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the straightforward generalization of augmented semi-simplicial sets and semi-cubical sets to an arbitrary cardinal, where morphisms are sequences of elements of an arbitrary set $\nu$, extended with $\kstar$. That is:

@@

\section{Our setting}

In this

\subsection{Working with sets in type theory}
Martin-L\"of's Type theory~\cite{martinlof84,?} is a logical formalism based on the notion of $\Type$ rather than \Set. It is the core of several tools for the formalization of mathematics such as Coq~\cite{} or Lean~\cite{}. Type Theory is a flexible formalism supporting different models. One of this model interprets types as topological spaces, where equality is interpreted as path, and substitutivity of equality as transport, leading to the development of Homotopy Type Theory~\cite{hottbook}, a foundation of mathematics alternative to set theory. Conversely, a notion of sets can be recovered within Type Theory as \U, denoting the subtype of $\Type$ for which paths are degenerated, what can be expressed by the property of Uniqueness of Identity Proofs (UIP). Technically, this is expressed as a structure equipping a domain $\Dom$ with the property $\UIP$:

@@

In the $\U$ universe, the following properties hold:

\begin{enumerate}
  \item[(i)] UIP holds on the unit and bool type:
\begin{align*}
  \textsf{eq\_refl}        & : \forall x, x = x                                                    \\
  \textsf{UIP\_refl\_unit} & : \forall u : \textsf{unit}, \forall x : u = u, x = \textsf{eq\_refl} \\
  \textsf{UIP\_refl\_bool} & : \forall b : \textsf{bool}, \forall x : b = b, x = \textsf{eq\_refl} \\
\end{align*}
  \item[(ii)] UIP propagates to $\Sigma$-types.
  \item[(iii)] UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

Note that an equivalent definition of UIP is given by \textsf{UIP\_refl}:

\begin{align*}
  \textsf{UIP\_refl} & : \forall x:\Dom, \forall p : x = x, p = \textsf{eq\_refl } x
\end{align*}

\subsection{Fibered versus indexed representation}
It is well-known in set theory that families of small sets can be represented in a fibered or indexed way. In the fibered way, a family over $T$ is represented a pair of a set $S$, and a function from $S$ to $T$. The function classifies the original set into fibers. In contrast, in the indexed way, a family over $T$ is represented as a functional relation with domain $T$, which itself, by the replacement axiom, is a set.

@@

The latter correspondence can be expressed in Homotopy Type Theory, for $T: \Type$, as an equivalence between $\Sigma S: \Type\;.\;S \rightarrow T$ and $T \rightarrow \Type$, which can be proved using univalence. This correspondence also holds on \U, for $T: \U$:

@@

Let us now come back to semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.

\begin{equation*}
  \begin{tikzcd}
    X_0 : \U & X_1 : \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2 : \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar R}" description, shift right=2] \arrow[l, "\delta^{\kstar L}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

@@

\section{Our construction}

In this section, we gradually build up an intuition for our construction, followed by the full details.

\subsection{Intuition for our formal construction}

\begin{notation}[$\Sigma$]
  In the prose that follows, we assume the existence of $\Sigma$-types with projections $\hd$ and $\tl$.
\end{notation}
